\chapter{Apptainer (formally known as Singularity)}
\label{ch:apptainer}

\section{What is Apptainer?}

Apptainer is an open-source computer program that performs operating-system-level virtualization (also known as containerisation).

One of the main uses of Apptainer is to bring containers and reproducibility to
scientific computing and the high-performance computing (HPC) world.
Using Apptainer/Singularity containers, developers can work in reproducible environments of their choosing and design,
and these complete environments can easily be copied and executed on other platforms.

For more general information about the use of Apptainer, please see the official documentation at \url{https://apptainer.org/docs/}.

This documentation only covers aspects of using Apptainer on the \hpcInfra infrastructure.

\section{Restrictions on image location}

Some restrictions have been put in place on the use of Apptainer. This is mainly done
for performance reasons and to avoid that the use of Apptainer impacts other users on the system.

The Apptainer/Singularity image file must be located on either one of the scratch filesystems,
the local disk of the workernode you are using or \lstinline|/dev/shm|. The centrally provided \lstinline|apptainer|
command will refuse to run using images that are located elsewhere, in particular on the \lstinline|$VSC_HOME|,
\lstinline|/apps| or \lstinline|$VSC_DATA| filesystems.

In addition, this implies that running containers images provided via a URL
(e.g., \lstinline|shub://...| or \lstinline|docker://...|) will not work.

If these limitations are a problem for you, please let us know via \hpcinfo.

\section{Available filesystems}

All HPC-UGent shared filesystems will be readily available in a Apptainer/Singularity container,
including the home, data and scratch filesystems, and they will be accessible via the
familiar \lstinline|$VSC_HOME|, \lstinline|$VSC_DATA*| and \lstinline|$VSC_SCRATCH*| environment variables.

\section{Singularity Images}

\subsection{Creating Singularity images}

Creating new Apptainer/Singularity images or converting Docker images, by default, requires admin privileges,
which is obviously not available on the \hpcInfra infrastructure. However,
if you use the \lstinline|--fakeroot| option, you can make new Apptainer/Singularity images or convert 
Docker images. 

Due to the nature of \lstinline|--fakeroot| option, we recommend to write your apptainer 
image to a globally writable location, like \lstinline|/tmp|, or \lstinline|/local| directories. 
Once the images is created, you should move it to your desired destination.
An example to make a apptainer container:
\begin{prompt}
%\shellcmd{APPTAINER_CACHEDIR=/tmp/ APPTAINER_TMPDIR=/tmp/ apptainer build --fakeroot /tmp/tensorflow-21.10-tf1-py3.sif docker://nvcr.io/nvidia/tensorflow:21.10-tf1-py3}%
\end{prompt}

\subsection{Converting Docker images}

For more information on converting existing Docker images to Apptainer/Singularity images,
see \url{https://apptainer.org/docs/user/main/docker_and_oci.html}.

We strongly recommend the use of Docker Hub, see \url{https://hub.docker.com/} for more information.

\section{Execute our own script within our container}

Copy testing image from \lstinline|/apps/gent/tutorials/Singularity| to \lstinline|$VSC_SCRATCH|:

\begin{prompt}
%\shellcmd{cp /apps/gent/tutorials/Singularity/CentOS7\_EasyBuild.img \$VSC\_SCRATCH/}%
\end{prompt}

Create a job script like:

\begin{code}{bash}
#!/bin/sh

#PBS -o apptainer.output
#PBS -e apptainer.error
#PBS -l nodes=1:ppn=1
#PBS -l walltime=12:00:00


apptainer exec $VSC_SCRATCH/CentOS7_EasyBuild.img ~/my_script.sh
\end{code}

Create an example \lstinline|myscript.sh|:

\begin{code}{bash}
#!/bin/bash

# prime factors
factor 1234567
\end{code}

\section{Tensorflow example}

We already have a Tensorflow example image, but you can also convert the Docker
image (see \url{https://hub.docker.com/r/tensorflow/tensorflow}) to a Apptainer/Singularity image yourself

Copy testing image from \lstinline|/apps/gent/tutorials| to \lstinline|$VSC_SCRATCH|:

\begin{prompt}
%\shellcmd{cp /apps/gent/tutorials/Singularity/Ubuntu14.04\_tensorflow.img \$VSC\_SCRATCH/}%
\end{prompt}

\begin{code}{bash}
#!/bin/sh
#
#
#PBS -o tensorflow.output
#PBS -e tensorflow.error
#PBS -l nodes=1:ppn=4
#PBS -l walltime=12:00:00
#

apptainer exec $VSC_SCRATCH/Ubuntu14.04_tensorflow.img python ~/linear_regression.py
\end{code}

You can download \lstinline|linear_regression.py| from
\href{https://github.com/tensorflow/tensorflow/blob/r1.12/tensorflow/examples/get_started/regression/linear_regression.py}{the official Tensorflow repository}.

\section{MPI example}

It is also possible to execute MPI jobs within a container, but the following requirements apply:

\begin{itemize}
    \item Mellanox IB libraries must be available from the container (install the \lstinline|infiniband-diags|, \lstinline|libmlx5-1| and \lstinline|libmlx4-1| OS packages)
    \item Use modules within the container (install the \lstinline|environment-modules| or \lstinline|lmod| package in your container)
    \item Load the required module(s) before \lstinline|apptainer| execution.
    \item Set \lstinline|C_INCLUDE_PATH| variable in your container if it is required during
        compilation time (\lstinline|export C_INCLUDE_PATH=/usr/include/x86_64-linux-gnu/:$C_INCLUDE_PATH| for Debian flavours)
\end{itemize}

Copy the testing image from \lstinline|/apps/gent/tutorials/Singularity| to \lstinline|$VSC_SCRATCH|

\begin{prompt}
%\shellcmd{cp /apps/gent/tutorials/Singularity/Debian8\_UGentMPI.img \$VSC\_SCRATCH/}%
\end{prompt}

For example to compile an
\href{https://github.com/open-mpi/ompi/blob/master/examples/ring_c.c}{MPI example}:

\begin{prompt}
%\shellcmd{module load intel}%
%\shellcmd{apptainer shell \$VSC\_SCRATCH/Debian8\_UGentMPI.img}%
%\shellcmd{export LANG=C}%
%\shellcmd{export C\_INCLUDE\_PATH=/usr/include/x86\_64-linux-gnu/:\$C\_INCLUDE\_PATH}%
%\shellcmd{mpiicc ompi/examples/ring\_c.c -o ring\_debian}%
%\shellcmd{exit}%
\end{prompt}

Example MPI job script:

\begin{code}{bash}
#!/bin/sh

#PBS -N mpi
#PBS -o apptainermpi.output
#PBS -e apptainermpi.error
#PBS -l nodes=2:ppn=15
#PBS -l walltime=12:00:00

module load intel vsc-mympirun
mympirun --impi-fallback apptainer exec $VSC_SCRATCH/Debian8_UGentMPI.img ~/ring_debian
\end{code}
